---
title: "Project 2"
author: "Diva Medina Camp, Wesley Newcomb."
date: "2023-05-20<br><br>"
output:
  html_document: default
  pdf_document: default
---

<style type="text/css">

h1.title {
  font-size: 38px;
  text-align: center;
}
h4.author { /* Header 4 - and the author and data headers use this too  */
    font-size: 18px;
  text-align: center;
}
h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  text-align: center;
}
<br>
</style>



```{r libraries, include=FALSE}
library(dplyr)
library(leaflet)
library(knitr)
#library(kableExtra)
library(tinytex)
```

### Objective 1 - Global Map

You are tasked to create a world map to gain an appreciation for where the most occurrences of COVID-19 confirmations and deaths are located.

Create this map using leaflet for the most recent date as shown below. For this map, sum the confirmations and deaths of provinces into one value to depict the total number for the country they belong to. When creating a marker for each country in the map, calculate lat and long as the mean values for the provinces that make up each country.

Customize the map to reflect the differences in magnitude for confirmations and deaths. In the example map below, circle markers that are blue represent low values, gray represents neutral values, and red represents high values. Low, middle, and high values were categorized to aesthetically map the markers based on their probabilistic distribution using the quartile function. You may use any method you like so that it is logical and allows visualization of value intensity. As well, customize the map to include hover labels that indicate country names and popup labels to show the value of confirmations and deaths for that country.

```{r map}

#extract data of global confirmed COVID cases
confirmed_COVID_global <- 
  read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv", stringsAsFactors=FALSE)

#extract data of global COVID deaths
deaths_COVID_global <- 
  read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv", stringsAsFactors=FALSE)

#variable holding dataframe with necessary columns for confirmed
lastday_confirmed <- 
  select(confirmed_COVID_global, Province.State, Country.Region, Lat, Long, "Latest_Day"=tail(names(confirmed_COVID_global),1))

#variable holding dataframe with necessary columns for deaths
lastday_deaths <-
  select(deaths_COVID_global, Province.State, Country.Region, Lat, Long, "Latest_Day"=tail(names(deaths_COVID_global),1))

#Create dataframe containing sum of deaths and confirmations by province
COVID_dataSum <- 
  lastday_confirmed %>% dplyr::inner_join(lastday_deaths, by=c("Province.State", "Country.Region", "Lat", "Long")) %>%
  dplyr::mutate(All_Latest = rowSums(across(c(Latest_Day.x, Latest_Day.y)))) %>%
  dplyr::rename("Last_Confirm"="Latest_Day.x", "Last_Deaths"="Latest_Day.y") 
  
#remove unnecessary data entries that involve Olympics or unknow origin
COVID_dataSum <- COVID_dataSum[-c(107, 245, 286),]

#Create dataframe which averages lats and longs, and adds confirmations and deaths.
COVID_dataSum_country <- 
  group_by(COVID_dataSum, Country.Region) %>%
  dplyr::summarize(mean(Lat,na.rm=TRUE), mean(Long,na.rm=TRUE), sum(Last_Confirm), sum(Last_Deaths), sum(All_Latest)) %>% 
  dplyr::rename("lat"=2, 
         "long"=3, 
         "Confirmations"=4, 
         "Deaths"=5, 
         "All"=6)
  
#format numeric data to have commas
COVID_dataSum_country$Confirmations <- prettyNum(COVID_dataSum_country$Confirmations, big.mark=",", scientific = FALSE)
COVID_dataSum_country$Deaths <- prettyNum(COVID_dataSum_country$Deaths, big.mark=",", scientific = FALSE)

#color palette function
pal = colorQuantile(palette=c("dodgerblue3", "grey35", "red"), domain=COVID_dataSum_country$All, n=5)

#Generate Map 
leaflet(COVID_dataSum_country) %>%
  addTiles() %>%
  setView(lng=0, lat=0, zoom=1) %>%
  addCircleMarkers(lng=~long, lat=~lat, radius=4, 
                   label=~Country.Region, color=~pal(COVID_dataSum_country$All), 
                   group="Confirmations",
                   popup=paste("Confirmed: ", as.character(COVID_dataSum_country$Confirmations))) %>%
  addCircleMarkers(lng=~long, lat=~lat, radius=4, 
                   label=~Country.Region, color=~pal(COVID_dataSum_country$All), 
                   group="Deaths",
                   popup=paste("Deaths: ", as.character(COVID_dataSum_country$Deaths))) %>%
  addLayersControl(overlayGroups = c("Confirmations", "Deaths"),
                   options = layersControlOptions(collapsed = FALSE))

```

### Objective 2 - Narrowing Down Hot Spots

Seeing the global map of COVID-19 cases results in the stark realization that some countries are more affected than others. In order to narrow down your studies, create a table using kable from knitr listing the top countries as far as confirmations and deaths (sum values for provinces of the same country into one value and show the country only). 


```{r data_prep}

#library(kableExtra)
library(tinytex)

#extract data of global confirmed COVID cases
confirmed_COVID_global <- 
  read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv", stringsAsFactors=FALSE)

#extract data of global COVID deaths
deaths_COVID_global <- 
  read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv", stringsAsFactors=FALSE)

#variable holding dataframe with necessary columns for confirmed
lastday_confirmed <- 
  select(confirmed_COVID_global, Province.State, Country.Region, 
         "Latest_Day"=tail(names(confirmed_COVID_global),1))

lastday_confirmed <- lastday_confirmed[-c(107 ,245 ,286),]

#variable holding dataframe with necessary columns for deaths
lastday_deaths <-
  select(deaths_COVID_global, Province.State, Country.Region, 
         "Latest_Day"=tail(names(deaths_COVID_global),1))

#remove unnecesary data points like Olympics and cruise ship.
lastday_deaths <- lastday_deaths[-c(107 ,245 ,286),]

#Make dataframe for confirmations data. Group by country, sum, add commas to numbers.
COVID_confirmed <- lastday_confirmed %>%
  dplyr::group_by(Country.Region) %>%
  dplyr::summarize(sum(Latest_Day)) %>%
  dplyr::rename("Counts"="sum(Latest_Day)", "Country"="Country.Region") %>%
  dplyr::arrange(desc(Counts))

COVID_confirmed$Counts <- prettyNum(COVID_confirmed$Counts, big.mark = ',', scientific=FALSE)

#Make dataframe for deaths data. Group by country, sum, arrange in desc order, add commas to numbers.
COVID_deaths <- lastday_deaths %>%
  dplyr::group_by(Country.Region) %>%
  dplyr::summarize(sum(Latest_Day)) %>%
  dplyr::rename("Counts"="sum(Latest_Day)", "Country"="Country.Region") %>%
  dplyr::arrange(desc(Counts))

COVID_deaths$Counts <- prettyNum(COVID_deaths$Counts, big.mark = ',', scientific=FALSE)

#Add rank numbers to table, and order columns.
COVID_confirmed <- COVID_confirmed %>% dplyr::mutate(Rank = row_number()) 

col_order <- c(3,1,2)
COVID_confirmed <- COVID_confirmed[, col_order]

```

```{r table, eval=FALSE}
# Combine confirmations and deaths tables.
COVID_Table <- cbind(COVID_confirmed, COVID_deaths)

# Format table
COVID_Table %>%
  kbl(align=c(rep("r",1), rep("l",1), rep("r",1), rep("l",1), rep("r",1)), 
      caption = "Table of Top Countries") %>%
  kable_styling(bootstrap_options = "striped", "responsive",
                full_width = FALSE) %>%
  add_header_above(c(" " = 1, "Confirmations" = 2, "Deaths" = 2), font_size="large") %>%
  scroll_box(width = "800px", height = "550px")
```

### Objective 3 - Zooming Into Our State and California

After reading the top tables, you are stunned! The US overtakes every other country in terms of COVID-19 confirmations. As such, you are concerned about the state you live in and would like to understand how COVID-19 events have shaped the trajectory of the disease. Create two scatter plots to gain a better understanding. The first scatter plot should be California’s trajectory for confirmations. The second scatter plot should show California’s top three city trajectories for confirmations. You are interested in studying how the vaccine affected the number of confirmations. The Moderna vaccine was first available as an emergency use authorized (EUA) vaccine and required two shots spaced six weeks apart. Indicate on the plots the day the second dosage was given to those that received the first dosage the day Moderna was EUA (January 29th, 2021). As a diligent scientist that knows that new COVID variants have mutations in the spike protein (the region that the vaccine was developed for), you also want to study how confirmation rates change as new variants become the dominant infectious strain. Indicate on the plots when the delta and omicron variants became the dominant strain in California (May 11th, 2021 and November 26th, 2021 respectively). In the example below, the function plot_grid from the R package cowplot was to organize the graphs into a grid to more easily compare statewide vs top city plots.

```{r Covid-19_US, echo=TRUE, eval=FALSE}

library(knitr)
library(magrittr)
library(RCurl)
library(ggplot2)
library(tidyr)
library(stringr)
library(dplyr)
library(leaflet)
library(tinytex)

# Extract data of US confirmed COVID cases
confirmed_US <- 
  read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv")

# Extract data of US COVID deaths
deaths_US <- 
  read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv")


# Reuses earlier data frame for only CA confirmed cases by county
dfConfirmedCAChartData <- confirmed_US %>%
  
  # removes all columns with numeric values that aren't case counts
  select(-(`UID`)) %>%
  select(-(`code3`)) %>%
  select(-(`FIPS`)) %>%
  select(-(`Lat`)) %>%
  select(-(`Long_`)) %>%
  
  # retains only the California rows
  filter(`Province_State` == "California") %>%
  
  # retains only the rows for San Diego, Riverside, LA
  filter(`Admin2` %in%
           c("San Diego", "Riverside", "Los Angeles")) %>%
  
  # groups by County (aka Admin2)
  group_by(`Admin2`) %>%
  
  # totals all columns that are numeric
  summarise_if(is.numeric, sum) %>%

  # gathers data into long form
  gather(key=Date, value=`Confirmed Cases`, -`Admin2`) %>%
  
  # converts the date column into real dates for the chart
  mutate(`Date`=as.Date(`Date`, format = "%m/%d/%y"))


## Isolate CA confirmations
confirmed_US %>% select(-c(UID, iso2, iso3, code3, FIPS, Admin2, Country_Region, Lat, Long_, Combined_Key)) %>% filter(Province_State == "California") %>% data.frame() -> CA
CA %<>% select(-Province_State)
colnames(CA) <- gsub("X","",colnames(CA))
colnames(CA) <- as.Date(colnames(CA) , format = "%m.%d.%y")
CA_sum <- colSums(CA)
date <- colnames(CA)
ready <- data.frame(date,CA_sum)
ready$date <- as.Date(ready$date)

## Making graph for state
p <- ggplot(ready, aes(date,CA_sum)) + geom_point(color="darkblue") + theme_gray() + labs(x="", y="Confirmations", title = "COVID-19 confirmations in California")

## plotting vertical lines
p1 <- p + geom_vline(aes(xintercept=as.Date("2020-03-19")),color="green", linetype = "longdash") + geom_vline(aes(xintercept=as.Date("2021-01-29")),color="darkgreen", linetype = "longdash") + geom_vline(aes(xintercept=as.Date("2021-07-04")),color="darkred", linetype = "longdash") + geom_vline(aes(xintercept=as.Date("2021-12-20")),color="brown", linetype = "longdash")

## labeling vertical lines
p2 <- p1 + geom_text(aes(as.Date("2020-03-19"), 7000000, label ="SAH", hjust="right"))  + geom_text(aes(as.Date("2021-01-29"), 7000000, label ="EUA + 6 weeks", hjust="right")) + geom_text(aes(as.Date("2021-07-04"), 7000000, label ="Delta", hjust="right")) + geom_text(aes(as.Date("2021-12-20"), 7000000, label ="Omicron", hjust="right"))

## adding tick marks
p3 <- p2 + scale_x_continuous(breaks = c(as.Date("2020-03-19"), as.Date("2021-01-29"), as.Date("2021-07-04"), as.Date("2021-12-20")), labels = c("2020-03-19", "2020-01-29", "2020-07-04", "2021-12-20")) 


c1 <- ggplot(dfConfirmedCAChartData,
       aes(x = `Date`, y = `Confirmed Cases`,
           group = `Admin2`)) +
  #geom_line(aes(color=`Admin2`), size = 2) +     
  geom_line(aes(color=`Admin2`), size = 1) +
  ggtitle("") +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Date", y = "Cases", color = "")

## plotting vertical lines
c2 <- c1 + 
geom_vline(aes(xintercept=as.Date("2021-03-19")),color="darkgreen", linetype = "longdash") + geom_vline(aes(xintercept=as.Date("2021-07-04")),color="darkred", linetype = "longdash") + geom_vline(aes(xintercept=as.Date("2021-11-20")),color="brown", linetype = "longdash")

## labeling vertical lines
c3 <- c2 + geom_text(aes(as.Date("2020-03-19"), 2500000, label ="", hjust="right"))  + geom_text(aes(as.Date("2021-01-29"), 2500000, label ="", hjust="right")) + geom_text(aes(as.Date("2021-07-04"), 2500000, label ="", hjust="right")) + geom_text(aes(as.Date("2021-12-20"), 2500000, label ="", hjust="right"))

## adding tick marks
c4 <- c3 + scale_x_continuous(breaks = c(as.Date("2020-03-19"), as.Date("2021-01-29"), as.Date("2021-07-04"), as.Date("2021-12-20")), labels = c("2020-07", "2021-01", "2021-07", "2022-01")) 

# Print graph as plot grid
plot_grid(p3, c4, ncol = 1, align = "h")
```
Below is a chart showing the deaths and confirmed cases for the state of California over time. The green vertical line provides a visual representation of when the first shutdowns occurred on March 19, 2020.

Below is a chart showing the confirmed cases over time, only for the counties of Los Angeles, Riverside, and San Diego. The start 2020-01 provides a visual representation of when the first shutdowns occurred on March 19, 2020.


### Objective 4 - Digging Deeper

Although these plots do not tell the whole story, they are great for helping us determine where to look. Different cities may have different populations, population densities, cultural discrepancies, compliance, and city regulations to name a few. We will explore the role of population on these metrics using visualizations. Arrange two scatter plots using cowplot’s plot_grid to show the relationship between population and confirmed counts as well as death counts and confirmed counts. You will need to use a log transform on all variables to show such a relationship. Please consult the example below for an idea of what this may look like. From these graphs we can see that population greatly affects confirmations and deaths. This coincides with our plots above as Los Angeles's population is 301% greater than San Diego’s population and 406% greater than Riverside’s population!

```{r deeper, echo=TRUE, eval=FALSE}

library(readr)
library(tidyverse)
library(cowplot)

# Extract data of US confirmed COVID cases
confirmed_US <- 
  read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv")

# Extract data of US COVID deaths
deaths_US <- 
  read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv")


## Make death vs confirmation for CA
Population <- deaths_US$Population
City <- deaths_US$Admin2
Confirmation <- confirmed_US$'5/28/21'
State <- confirmed_US$Province_State
Deaths <- deaths_US$'5/28/21'

# Make plot
p1 <- ggplot(data = data) +  
  geom_point(mapping = aes(x = Population, y = Confirmation), color = "darkblue") +
  labs(
    title = "Confirmations Vs. Population",
    x = "Population",    
    y = "Confirmation Counts"
  ) + theme_bw() + 
  theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) +
  theme_minimal() +
  scale_y_continuous(trans = "log2", limits=c(16,8388608), labels = scales::comma) +  
  scale_x_continuous(trans = "log2", limits=c(128,4194304), labels = scales::comma) +  
  theme(legend.title = element_blank())

## Use plot_grid
p2 <- ggplot(data = data) +
  geom_point(mapping = aes(x = Confirmation, y = Deaths), color = "darkred") +
  labs(
    title = "Deaths Vs. Confirmations",
    x = "Confirmed Counts",
    y = "Deaths Counts"
    
  ) + theme_bw() + 
  theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) +
  theme_minimal() +
  scale_y_continuous(trans = "log2", limits=c(8,32768), labels = scales::comma) +  
  scale_x_continuous(trans = "log2", limits=c(128,4194304), labels = scales::comma) +  
  theme(legend.title = element_blank())

# Side by side
plot_grid(p1, p2)
```


##### CSIT-165,
###### 05/22/2023

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
